Manuscript for Foodborne-Outbreak Project

### 2023-04-22 code/analysis_code/exploratory_analysis.qmd
Weifan_final

### Bacterial Etiologies for Outbreaks
```{r}
# General
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
         filter(Pathogen_Type == "Bacteria")%>%
         group_by(Simplified_Etiology)%>%
         mutate(all_hospitalizations=sum(Hospitalizations)),
       aes(x = fct_reorder(Simplified_Etiology,all_hospitalizations))) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

@@

@@

### Food Type and Hospitalizations
```{r}
# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations), 
       aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
           )) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
  coord_flip() + theme(legend.position = "none")

>>>>

### Bacterial Etiologies for Outbreaks
```{r}
# General
## Hospitalizations
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
         filter(Pathogen_Type == "Bacteria")%>%
         group_by(Simplified_Etiology)%>%
         mutate(all_hospitalizations=sum(Hospitalizations)),
       aes(x = fct_reorder(Simplified_Etiology,all_hospitalizations))) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

@@

## Illnesses
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Illnesses) %>%
         filter(Pathogen_Type == "Bacteria")%>%
         group_by(Simplified_Etiology)%>%
         mutate(all_Illnesses=sum(Illnesses)),
       aes(x = fct_reorder(Simplified_Etiology,all_Illnesses))) + geom_bar() + 
  labs(x = "Bacteria", y = "Illnesses", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

ggsave("bacterial_etiology_v_Illnesses_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")

@@

### Food Type and Hospitalizations
```{r}
# Food Type and Hospitalizations
food_hosp=training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations)%>%
  arrange(desc(all_hospitalizations))
save_location <- here("results","tables", "food_v_hospitalizations.rds")
saveRDS(food_hosp,file = save_location)

  ggplot(food_hosp,aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
           )) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
  coord_flip() + theme(legend.position = "none")

### 2023-04-22 code/analysis_code/statistical_analysis.qmd
Weifan_final

```
According to the result, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2. However, none of those models can catch outliers (Illness > 1000).

@@

```
According to the results, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2 (14%). But again, outliers (Hospitalizations > 100) cannot be predicted well by any of those models.

@@

```{r}
# Illness
save_location <- here("results", "illness_rmse_models.rds")
saveRDS(illness_model_dataframe, file = save_location)

# Hospitalizations
save_location <- here("results", "hospital_rmse_models.rds")
saveRDS(hosp_model_dataframe, file = save_location)
```

>>>>

```
### Organize performances metrics for single variate models
```{r}

simp_illness_metric=lm_simple_resampled_illness%>%
    collect_metrics() %>%
  mutate(model="linear")%>%
  filter(.metric=='rmse')%>%
  bind_rows(poisson_simple_resampled_illness %>%
  collect_metrics() %>%
  filter(.metric=='rmse')%>%
    mutate(model="Poisson"))

save_location <- here("results", "tables","illness_simple_rmse.rds")
saveRDS(simp_illness_metric, file = save_location)
```

According to the result, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2. However, none of those models can catch outliers (Illness > 1000).

@@

### Organize performances metrics for single variate models
```{r}
simp_hospital_metric=lm_simple_resampled_hosp%>%
    collect_metrics() %>%
  filter(.metric=='rmse')%>%
  mutate(model="linear")%>%
  bind_rows(poisson_simple_resampled_hosp%>%
  collect_metrics() %>%
  filter(.metric=='rmse')%>%
    mutate(model="poisson"))

save_location <- here("results", "tables","hospital_simple_rmse.rds")
saveRDS(simp_hospital_metric, file = save_location)
```

According to the results, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2 (14%). But again, outliers (Hospitalizations > 100) cannot be predicted well by any of those models.

@@

```{r}
# Illness
# Univariate

# Multivariate
save_location <- here("results","tables", "illness_rmse_models.rds")
saveRDS(illness_model_dataframe, file = save_location)

# Hospitalizations
save_location <- here("results", "tables","hospital_rmse_models.rds")
saveRDS(hosp_model_dataframe, file = save_location)
```

### 2023-04-22 products/manuscript/Manuscript2.qmd
Weifan_final

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
output: word_document
bibliography: ../references.bib
csl: ../apa.csl
---

@@

@@

@@

After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations). We observed certain patterns between preditors and outcomes and selected

@@

After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set. In the United States, Multi-state outbreaks were the most commonly associated with hospitalizations caused by foodborne illnesses, followed by California, Ohio, and New York. Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2). Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2005, as shown in Table 2.
Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks. Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February.
Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3). Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5). In contrast, *E. coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier. 
Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6).
The testing data set revealed a number of trends that were consistent with those found in the training data set. For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8).
However, some shifts were also observed between the two data sets. While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9).
In addition, the testing data set revealed that *E. coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).

@@

Between univariate and multivariate regressions, multivariate models performed better with lower associated RMSEs. For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better with  and than multivariate poisson regression models `r $RMSE = 59.7, 5.64).

@@

```{r Model RMSE Comparison, paged.print=TRUE}
# Illness Models 
rmarkdown::paged_table(readRDS("../../results/illness_rmse_models.rds"))

# Hospitalization Models
rmarkdown::paged_table(readRDS("../../results/hospital_rmse_models.rds"))
```

Overall, between itself and the other models (linear regression, poisson regression, and bagging) generated, the random forest model performed the best for both health outcomes with the lowest RMSEs (55.7, 5.05).

@@

From 1998 to 2015, multistate outbreaks caused the most illnesses (N = 18,617) and hospitalizations (3,506) than any in-state outbreak. The fact that multistate outbreaks were seen more often than in any individual state for illnesses and hospitalizations suggests that these disease outbreaks typically involve a regional or national ingredient distributor. Hence, it would be reasonable to suggest laboratory sampling of goods, especially for bacteria, as a safety benchmark upon arrival at their destinations.

In terms of food categories, poultry was identified to have caused the most hospitalizations (N = 770), followed closely by fruits and beef. Interestingly, however, salads were found to cause the most illnesses (N = 11,493). Upon addition of total hospitalizations and illnesses for both poultry and salads, salads were determined to have the highest number of cases in total, which confirms our original hypothesis.

@@

The remaining two of the three initial hypotheses were proved incorrect. Instead of *E. coli*, results overall indicated that *Salmonella* spp. are the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data. Finally, we also see that food poisoning is most frequently linked to restaurants instead of private households.

\[Something about machine learning here???\]

@@

@@

```{r}
 first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
   rename(Hospitalizations = all_hospitalizations)
first_table <- flextable(first_table) 
first_table <- theme_vanilla(first_table) 
first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
first_table <- set_table_properties(first_table, layout = "autofit")
first_table
```

```{r}
second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
  select(`IFSAC Category`, all_illnesses) %>% 
  rename(Illnesses = all_illnesses)
second_table <- flextable(second_table)
second_table <- theme_vanilla(second_table)
second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
second_table
```

>>>>

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
output:
  bookdown::word_document2:
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

@@

```{r setup, include=FALSE, echo=FALSE, cache.lazy=FALSE}
# Setting options for the rest of the document
knitr::opts_chunk$set(warning=FALSE)

# Rounding rules
options(digits = 3)
#Load clean data from "processed_training_data.rds"
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)

#Load table from "state_v_hospitalization_training.rds"
# Path to Data
data_location <- here(
  "results","tables","state_v_hospitalization_training.rds")
# Import Data
state_hosp <- readRDS(data_location)

# Load table from "state_v_illness_training.rds"
# Path to Data
data_location <- here(
  "results","tables","state_v_illness_training.rds")
# Import Data
state_illness <- readRDS(data_location)

# Load table from "food_v_hospitalizations.rds"
# Path to Data
data_location<-here("results","tables", "food_v_hospitalizations.rds")
# Import Data
food_hosp<-readRDS(data_location)

# Load table from "training_food_vs_illnesses.rds"
# Path to Data
data_location<-here("results","tables", "training_food_vs_illnesses.rds")
# Import Data
food_illness<-readRDS(data_location)

# Load rmse data for outcome `Illnesses`

# Univariate for linear and poisson regression models
# Path to Data
data_location<-here("results","tables","illness_simple_rmse.rds")
# Import Data
illness_uni_rmse<-readRDS(data_location)

# Multivariate
# Path to Data
data_location=here("results","tables","illness_rmse_models.rds")
# Import Data
illness_rmse<-readRDS(data_location)

# Load rmse data for outcome `Hospitalizations`
# Univariate for linear and poisson regression models
# Path to Data
data_location<-here("results","tables","hospital_simple_rmse.rds")
# Import Data
hosp_uni_rmse<-readRDS(data_location)

# Multivariate
# Path to Data
data_location=here("results","tables","hospital_rmse_models.rds")
# Import Data
hosp_rmse<-readRDS(data_location)

# Final model metrics

```

@@

In the United States, it has been reported that one in six Americans acquire foodborne illnesses annually, leading to around 128,000 hospitalizations and 3,000 deaths.  Our project aims to identify the major etiology agents responsible for foodborne illnesses and their associated factors in the U.S. We utilized two datasets collected by the CDC, which contain foodborne outbreak information from 1998-2015 and 2009-2020, respectively. The first dataset was used for model training, while the second dataset was reserved for testing purposes. We developed regression models, including linear and Poisson regression, to predict health outcomes using selected predictors from the training dataset. Additionally, random forest model and bagging approaches were employed to refine the model. Results showed that multi-state outbreaks were the most commonly associated with total hospitalizations. Poultry was identified as the most frequent food vehicles linked to mass hospitalizations due to foodborne illnesses while salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015. Bacteria is the predominant etiologic agent that causes hospitalization, among which the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations. Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry. The best model constructed based on time, pathogen type, pathogen etiology, state, and food type was random forest model, with the lowest RMSE of `r illness_rmse$RMSE[1]`, `r hosp_rmse$RMSE[1]` for illnesses and hospitalizations, respectively.

@@

After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations). We observed certain patterns between predictors and outcomes and selected predictors accordingly for further model construction.

@@

After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set. The illnesses caused by a single outbreak ranged from `r min (training_data_final$Illnesses)` to `r max (training_data_final$Illnesses)` while resulted hospitalizations varied from `r min (training_data_final$Hospitalizations)` to `r max (training_data_final$Hospitalizations)` from 1998 to 2015. In the United States, Multi-state outbreaks were the most commonly associated with total hospitalizations (N= `r max (state_hosp$all_hospitalizations)`) caused by foodborne illnesses, followed by California (N=`r state_hosp$all_hospitalizations[2]`), Ohio (N=`r state_hosp$all_hospitalizations[3]`), and New York (n=`r state_hosp$all_hospitalizations[4]`). Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2). Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015, as shown in Table 2.
Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks. Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February.
Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3). Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5). In contrast, *E. coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier. 
Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6).
The testing data set revealed a number of trends that were consistent with those found in the training data set. For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8).
However, some shifts were also observed between the two data sets. While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9).
In addition, the testing data set revealed that *E. coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).

@@

Between univariate and multivariate regressions, multivariate models performed better with lower associated RMSEs. For both health outcomes, the multivariate linear regression models (RMSE = `r illness_rmse$RMSE[2]`, `r hosp_rmse$RMSE[3]`) constructed performed better with  and than multivariate poisson regression models (RMSE = `r illness_rmse$RMSE[3]`, `r hosp_rmse$RMSE[4]`).

@@

```{r Model RMSE Comparison, paged.print=TRUE}
# Illness Models 
rmarkdown::paged_table(readRDS("../../results/tables/illness_rmse_models.rds"))

# Hospitalization Models
rmarkdown::paged_table(readRDS("../../results/tables/hospital_rmse_models.rds"))
```

Overall, between itself and the other models (linear regression, poisson regression, and bagging) generated, the random forest model performed the best for both health outcomes with the lowest RMSEs (`r illness_rmse$RMSE[1]`, `r hosp_rmse$RMSE[1]`).

@@

From 1998 to 2015, multistate outbreaks caused the most illnesses (N = `r state_illness$all_illnesses[1]`) and hospitalizations (N = `r max (state_hosp$all_hospitalizations)`) than any in-state outbreak. The fact that multistate outbreaks were seen more often than in any individual state for illnesses and hospitalizations suggests that these disease outbreaks typically involve a regional or national ingredient distributor. Hence, it would be reasonable to suggest laboratory sampling of goods, especially for bacteria, as a safety benchmark upon arrival at their destinations.

In terms of food categories, poultry was identified to have caused the most hospitalizations (N = `r food_hosp$all_hospitalizations[1]`), followed closely by fruits (N = `r food_hosp$all_hospitalizations[2]`) and beef (N = `r food_hosp$all_hospitalizations[3]`). Interestingly, however, salads were found to cause the most illnesses (N = `r max(food_illness$all_illnesses)`). Upon addition of total hospitalizations and illnesses for both poultry and salads, salads were determined to have the highest number of cases in total, which confirms our original hypothesis.

@@

The remaining two of the three initial hypotheses were proved incorrect. Instead of *E. coli*, results overall indicated that *Salmonella* spp. were the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data. As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks (@jackson_outbreak-associated_2013). The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of  *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings (@skarzynska_distribution_2017).  The study revealed that *Salmonella* most frequently recovered from poultry samples, and *S*. Typhimurium being the most prevalent in pork products.

Finally, we observed that food poisoning is most frequently linked to restaurants instead of private households. This result is in agreement with a previous study that revealed over 40 % of outbreaks could be traced back to restaurants (@hedberg_systematic_2006). In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants (@li_surveillance_2020).

\[Something about machine learning here???\] #you can briefly talk about the algorithm or working principle behind random forest, what attributes may make it the best model?

@@

Future research can focus on the associations between subspecies of identified common bacterial etiology and crucial factors. Text mining approach might be implemented to explore more food carriers linked to foodborne outbreaks.

@@

```{r}
 first_table <- readRDS("../../results/tables/state_v_hospitalization_training.rds") %>% 
   rename(Hospitalizations = all_hospitalizations)
first_table <- flextable(first_table) 
first_table <- theme_vanilla(first_table) 
first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
first_table <- set_table_properties(first_table, layout = "autofit")
first_table
```

```{r}
second_table <- readRDS("../../results/tables/training_food_vs_illnesses.rds") %>% 
  select(`IFSAC Category`, all_illnesses) %>% 
  rename(Illnesses = all_illnesses)
second_table <- flextable(second_table)
second_table <- theme_vanilla(second_table)
second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
second_table
```
